---
title: "Getting Started with sobol"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sobol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **sobol** package generates low‑discrepancy Sobol sequences and is designed from the ground up for **parameter space exploration**.  
This tutorial will take you from the simplest possible usage to advanced reproducibility and parallel workflows.

## Installation

```{r, eval = FALSE}
# From GitHub
devtools::install_github("alrobles/sobol")
```

```{r}
library(sobol)
```

## 1. The quickest path: `sobol_design()`

Imagine you are tuning a machine‑learning model with three hyperparameters:

- `learning rate` in `[0.0001, 0.1]`
- `momentum` in `[0, 0.99]`
- `dropout` in `[0, 0.5]`

You want to explore the space with 200 well‑spread points. `sobol_design()` returns a **data frame** ready to be fed into your objective function.

```{r}
design <- sobol_design(
  lower = c(learning_rate = 0.0001, momentum = 0.00, dropout = 0.0),
  upper = c(learning_rate = 0.1000, momentum = 0.99, dropout = 0.5),
  nseq   = 200
)

head(design)
```

Points are in the exact ranges you specified:

```{r}
summary(design)
```

The design is **deterministic** and **space‑filling** – already a big improvement over simple random or grid search.

```{r, eval = FALSE}
# Use the design directly inside your optimisation loop
results <- purrr::pmap_dbl(design, ~ my_model(lr = ..1, mom = ..2, drop = ..3))
```

## 2. What makes `sobol_design` special?

Behind the scenes it calls `sobol_points()` to generate a **Sobol sequence** in the unit cube and then scales each column to your bounds.  

* **Low discrepancy** – points are more evenly distributed than random.  
* **Reproducible** – you’ll get the exact same design every time.  
* **No wasted points** – every evaluation adds information about the shape of your function.

Try it against a random Latin hypercube:

```{r, eval = FALSE}
# Not run, but you can compare visual uniformity
plot(design$learning_rate, design$momentum, col = "steelblue",
     main = "Sobol design (200 points)")
```

## 3. Going one level deeper: raw points

If you already have your own scaling logic, or need the raw `[0,1)` points, use `sobol_points()` directly.

```{r}
raw <- sobol_points(n = 512, dim = 4)
dim(raw)           # 512 rows, 4 columns
range(raw)         # values in [0, 1)
```

`sobol_points()` accepts an optional `skip` argument that lets you start from an arbitrary index – perfect for parallel workers (see below).

## 4. Incremental generation with `sobol_generator()`

Sometimes you don’t know in advance how many points you’ll need. Maybe you want to **evaluate a few, check convergence, then generate more**. That’s where the stateful generator shines.

```{r}
gen <- sobol_generator(dim = 3)

# Generate one point
sobol_next(gen)

# Generate a batch of 50
batch <- sobol_next_n(gen, n = 50)
dim(batch)  # 50 x 3

# What’s the current index?
sobol_index(gen)
```

You can also **jump** to any position:

```{r}
sobol_skip_to(gen, 1000)
sobol_index(gen)
```

This is the key to **parallel** and **restart‑friendly** workflows.

## 5. Reproducibility and parallel optimisation

All sequences are deterministic. So two calls with the same parameters will always match:

```{r}
a <- sobol_design(lower = c(p = 0), upper = c(p = 1), nseq = 32)
b <- sobol_design(lower = c(p = 0), upper = c(p = 1), nseq = 32)
identical(a, b)   # TRUE
```

To distribute work across multiple cores or machines, assign each a **non‑overlapping skip interval**.  

* Worker 1: points `0 – 999`  
* Worker 2: points `1000 – 1999`  
* Worker 3: points `2000 – 2999`

```{r}
# Worker 1
w1 <- sobol_design(lower = c(lr = 0.0001, mom = 0, drop = 0),
                   upper = c(lr = 0.1,    mom = 0.99, drop = 0.5),
                   nseq = 1000)  # implicitly starts at 0

# Worker 2 (needs raw points + skip to 1000)
raw2 <- sobol_points(n = 1000, dim = 3, skip = 1000)
# Then scale raw2 manually, or use sobol_design in the future with a skip argument
```

*(A `skip` argument for `sobol_design()` is under consideration – once available, parallel designs become one‑liners.)*

## 6. Advanced: chaining generators for adaptive sampling

A generator can be “rewound” at any time to re‑evaluate a segment:

```{r}
gen <- sobol_generator(dim = 2)
first_10 <- sobol_next_n(gen, n = 10)

# Oops, need to re‑evaluate the first 10 with different parameters
sobol_skip_to(gen, 0)
replicated <- sobol_next_n(gen, n = 10)
identical(first_10, replicated)  # TRUE
```

## 7. Performance notes

The C++ engine is heavily optimised. Even 1 000 000 points in 10 dimensions complete in under a second on modern hardware, freeing you to spend time on your actual model.

For extremely high dimensions (>1000) the engine falls back to runtime generation – still fast, but initialisation takes a tick longer. Precomputed tables cover the first 1000 dimensions instantly.

## Next steps

- **Reference** – `?sobol_design`, `?sobol_points`, `?sobol_generator`  
- **Articles** – “Performance Optimisation” and “Testing Strategy” on the package website  
- **Real‑world example** – check the `inst/examples/usage_examples.R` file  

```{r}
# Clean up
rm(design, raw, gen, a, b, first_10, replicated)
```

---

That’s all you need to start exploring your parameter space smarter and faster.  
**Welcome to `sobol`!**
